---
title: "Lab 3 Answer Sheet"
author: "BIOS 10602 - Multiscale Modeling of Biological Systems"
output: html_document
---

```{r eval = FALSE}
install.packages('seqinr', repos='http://cran.us.r-project.org')
```

```{r echo = FALSE, message = FALSE}
#loads the "seqinr" library into the R Markdown environment
library("seqinr")
require("seqinr")
```

## Part 1 - Sequence Analysis

```{r}
covid <- read.fasta(file="covid19.fasta")
covid_seq <- covid[[1]]
e_coli <- read.fasta(file="covid19.fasta")
e_coli_seq <- e_coli[[1]]
```


1.1
```{r}
myvec <- 3:11
myvec[4] <- "Hi"
print (myvec)
```
<span style="color:blue"> When a string is added to a integer vector, all the elements in the vector are converted to strings, as a vector can only have one type of value. </span>

```{r}
motleyCrew <- list(char = c("a", "b", "c"), num = 1:7, func = sample, logic = TRUE, missing = NA, numberList = list(one = 1, two = 2, three = 3))
```

1.2
```{r}
listTypes <- function(l) {
  for (item in l) {
    print (class(item))
  }
}

listTypes(motleyCrew)
```

```{r}
covid_seq <- covid[[1]]
```

1.3
```{r}
length(e_coli_seq)
```

1.4
```{r}
l <- length(e_coli_seq)
e_coli_seq[(l-25):l]
```

## Part 2: Basic DNA analysis

### Complementary Strand

```{r}
init_strand <- c("c", "g", "a", "g", "a", "c", "a")
```

2.1
```{r}
compStrand <- function(seq) {
  switchNT <- function(nt) {
    if (nt == "c") return ("g")
    if (nt == "g") return ("c")
    if (nt == "t") return ("a")
    if (nt == "a") return ("t")
  }
  
  comp <- numeric(length(seq))
  
  for (i in 1:length(seq)) {
    comp[length(seq) - i + 1] <- switchNT(seq[i])
  }
  
  return (comp)
}
```

### Counts and Frequencies

2.2
```{r}
table(e_coli_seq)
```
<span style="color:blue"> a: 1141382, c: 1177437, g:1180091, t: 1142742 </span>

2.3
```{r}
frequencyAdenine <- function(seq) {
  return (table(seq)["a"]/length(seq))
}

frequencyThymine <- function(seq) {
  return (table(seq)["t"]/length(seq))
}
```

2.4 a
<span style="color:blue"> The ribonucleotide monophosphates corresponding to A, C, and G have an additional hydroxyl group where a hydrogen atom would be in deoxyribonucleotide monophosphates. As such, the ribonucleotide A, C, and G are about an oxygen atom heavier (~16 amu) than their deoxyribonucleotide counterparts. U and T, uracil and thymine, are different molecules, as thymine has an additional H3C. </span>

2.4 b
```{r}
seqWeight <- function(seq, type="DNA") {
  dnaWeight <- function(nt) {
    if (nt == "c") return (289.2)
    if (nt == "g") return (329.2)
    if (nt == "t") return (304.2)
    if (nt == "a") return (313.2)
  }
  
  rnaWeight <- function(nt) {
    if (nt == "c") return (305.2)
    if (nt == "g") return (345.2)
    if (nt == "u") return (306.2)
    if (nt == "a") return (329.2)
  }
  
  if (type == "DNA") {
    sum <- 0
    for (nt in seq) {
      sum <- sum + dnaWeight(nt)
    }
    return (sum)
  }
  
  sum <- 0
  for (nt in seq) {
    sum <- sum + rnaWeight(nt)
  }
  return (sum)
}
```

```{r}
seq <- c("g", "g", "t", "a", "t", "t", "c", "c", "g", "a", "g", "a", "c", "a")
```

2.4 c
```{r}
seqWeight(seq)
```


### GC Content of DNA

2.5
```{r}
GC_Content <- function(seq) {
  counts <- table(seq)
  return ((counts["g"] + counts["c"])/length(seq))
}
GC_Content(e_coli_seq)
```
<span style="color:blue"> Yes </span>

2.6
```{r}
windowStartPositions <- function(length, size, step) {
  return (seq(1, length - size + 1, step))
}
```

2.7
```{r}
slidingGC <- function(seq, size, step) {
  windows <- windowStartPositions(length(seq), size, step)
  
  gc <- length(windows)
  
  for (i in 1:length(windows)) {
    gc[i] <- GC_Content(seq[windows[i]:(windows[i]+size-1)])
  }
  
  return (gc)
}
```

2.8
```{r}
plotSlidingGC <- function(seq, size, step) {
  plot(x=windowStartPositions(length(seq), size, step), y=slidingGC(seq, size, step), pch=".", type="o", main="GC Content versus Sequence Position", xlab="Sequence Window Position", ylab="GC Content")
}
```
<span style="color:blue"> Decreasing the size and step both result in closer, more volatile data points. Decreasing just size allows for detection of greater changes in GC content that may occur in smaller stretches of the sequence. Increasing generally has the opposite effects than those described above. </span>

## Part 3: Finding the origin of DNA replication (*oriC*)

### Finding *oriC*

```{r}
E_Coli <- read.fasta("E_coli.fasta")[[1]]
t_petrophila <- read.fasta("Thermotoga petrophila.fasta")[[1]]
s_solfataricus <- read.fasta("Sulfolobus solfataricus.fasta")[[1]]
```

3.1 a
```{r}
skew <- function(seq, i) {
  t <- table(seq[1:i])
  return (t["g"] - t["c"])
}
```

3.1 b
```{r}

skewDiagram <- function(seq) {
    skew <- 0
    
    skews <- numeric(length(seq))
    
    for (i in 1:length(seq)) {
      if (seq[i] == "g") {
        skew <- skew + 1
      } else if (seq[i] == "c") {
        skew <- skew - 1
      }
      skews[i] <- skew
    }
  
  plot(x=seq(1, length(skews), 8000), y=skews[seq(1, length(skews), 8000)], main="Skew Diagram", ylab="Skew")
}

```

3.1 c
```{r}
min_skew <- function(seq) {
  skew <- 0
  
  minskew <- Inf
  min <- 0
  
  for (i in 1:length(seq)) {
    if (seq[i] == "g") {
      skew <- skew + 1
    } else if (seq[i] == "c") {
      skew <- skew - 1
    }
    
    if (skew < minskew) {
      minskew <- skew
      min <- i
    }
  }
  
  return (min)
}
```

3.1 d
```{r}
skewDiagram(s_solfataricus)
min_skew(s_solfataricus)
```
<span style="color:blue"> The origin of replication is most likely to be found around nucleotide 749375. This nucleotide is where the skew, or the difference between guanine and cytosine counts is greatest when calculated cumulutively up to that point. It is also visually among the most jarring and immediate drops in skew, when referencing the skew diagram. 

The other, smaller, yet still significant valleys could also indicate origins of replication. In fact, it may indicate there were multiple sites of replication. 

See https://pubmed.ncbi.nlm.nih.gov/15107501/
</span>

3.2
```{r}
skewDiagram(t_petrophila)
```
<span style="color:blue"> The most apparent dip in skew is around nucleotide 800,000. The minimum skew corroborates this.  </span>

3.3
```{r}
skewDiagram <- function(seq) {
    skew <- 0
    
    skews <- numeric(length(seq))
    
    for (i in 1:length(seq)) {
      if (seq[i] == "g") {
        skew <- skew + 1
      } else if (seq[i] == "c") {
        skew <- skew - 1
      }
      skews[i] <- skew
    }
  
  plot(x=seq(1, length(skews), 100), y=skews[seq(1, length(skews), 100)], main="Skew Diagram", ylab="Skew")
}

skewDiagram(covid_seq)
```